Coursera - Reproducible Research - Peer Assessment 1  
====================================================  

This is the R Markdown file for the first peer assessment assignment.  
The data file must be downloaded, unzipped, then loaded from its CSV  
format into a data frame.  

First download, unzip, and load  
-------------------------------  
We download the zipped file from the URL given in the assignment page.  
We then unzip the downloaded file into its uncompressed csv format.  
We load the data from the csv format file into our working data frame.  

```{r}
zip.file <- download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", "repdata_data_activity.zip", method="curl")
file.conn <- unz("repdata_data_activity.zip", "activity.csv")
activ.dat <- read.csv("activity.csv", header=TRUE, sep=",", na.strings="NA")
```

Create directory ./figures/ in order to save plots and figures  
-------------------------------------------------------------- 

NOTE: the instructions in the assignment require a "./figures" directory.  
NOTE: the knitter directly created a a "./figure" directory (notice no "s").  
So, given the confusion on names (figure vs figures), I decided to plot twice.  
So, you will see that each plot was performed on the standard graphic device.  
This results in the plots that go under the "./figure" directory (with no s).  
You will also see that each plot was re-performed into a named png file.  
I placed this second named png file into the "./figures" directory.  

All this a little clumsy, but the instructions had that little bug in them...  

```{r}
if (!file.exists("./figures"))
    dir.create("./figures")
```

Histogram, mean and median of total steps per day  
-------------------------------------------------  
We create a subset data frame by removing the rows where the steps are NA.  
We then calculate the daily sums of steps by using the date as the factor  
along whose levels the sums are calculated, and store these in a vector.  
We then remove from this vector of sums the entires that are NA.  
These would be due to left-over dates where all initial steps were NA.  

```{r}
# Remove rows where there are NA values for steps in any interval

activ.dat.no.na <- activ.dat[!is.na(activ.dat$steps),]

# Sum the steps within each level of the factor "date", to get the daily sums

daily.sums <- tapply(activ.dat.no.na$steps, activ.dat.no.na$date, sum, simplify=T)

# Remove the NA values

daily.sums <- subset(daily.sums, !is.na(daily.sums))

# Plot histogram of daily steps on screen

noquote("Plotting Histogram of Daily Steps")
hist(daily.sums)

# Plot histoggram of daily steps in a png file

png(file="./figures/first-question-daily-steps-histogram.png")
hist(daily.sums)
dev.off()

# Calculate mean and print it

daily.mean <- mean(daily.sums)
noquote("Daily Step Mean is: ")
daily.mean

# Calculate median and print it

daily.median <-median(daily.sums)
noquote("Daily Step Median is: ")
daily.median
```

Plot, mean and median of steps per 5-minute interval, averaged over all days  
----------------------------------------------------------------------------  

Rather than looking at the data by the day factor, we now analyse it by interval.  
So we average over all days, the steps that were taken in each of the 5-min intervals.  

```{r}

# Use tapply to apply the mean along the levels of the interval viewed as a factor

interval.means <- tapply(activ.dat.no.na$steps, activ.dat.no.na$interval, mean, simplify=T)

# Now plot the values as requested in the assignment:
# First plot without the x-axis, in order to add better formatted times
# Then add the x-axis using the axis() command

plot(levels(as.factor(activ.dat.no.na$interval)), interval.means, xaxt = "n", xlab="5-Min Intervals", ylab="Average Steps", main="Line Plot of Average Steps Versus 5-Min Intervals", type="l")

# Create a vector of labels to mark the ticks on the x-axis
# Put these labels at the computed locations for tick marks
# Parameter 1 is what describes the x-axis in axis-related functions
# Also, draw a vertical blue line to mark the interval with max steps

x.labels <- c("00:00", "05:00", "10:00", "15:00", "20:00")
axis(1, at=axTicks(1), labels=x.labels)
abline(v=activ.dat.no.na$interval[which.max(interval.means)], col="blue")

# Now redo the same plotting into a png file

png(file="./figures/average-steps-per-5min-intervals-line-plot.png")
plot(levels(as.factor(activ.dat.no.na$interval)), interval.means, xaxt = "n", xlab="5-Min Intervals", ylab="Average Steps", main="Line Plot of Average Steps Versus 5-Min Intervals", type="l")
axis(1, at=axTicks(1), labels=x.labels)
abline(v=activ.dat.no.na$interval[which.max(interval.means)], col="blue")
dev.off()

# Now print the value of the interval with the largest average steps
# Mark the starting and ending times of that interval, to show both

int.start <- as.integer(activ.dat.no.na$interval[which.max(interval.means)])

# Calculate end of the interval
# Most of the time, this is 5 plus the start of the interval
# Except when the start is some hour plus 55 minutes...
# ... in which case the end interval is the next hour...
# ... which gets written as the interval start plus 45.
# Example: 155 refers to 01:55, and the next 5-min boundary is 200 for 02:00

if (int.start %% 100 == 55) {
    int.end <- int.start + 45
} else {
        int.end <- int.start + 5
}
int.start <- as.character(int.start)
int.end <- as.character(int.end)

# Stuff with leading zeros and insert ":" to pretty things up

pretty <- function (string) {
    switch (nchar(string), paste("000", string, sep=""), paste ("00", string, sep=""), paste("0", string, sep=""), string)
}
int.start <- pretty(int.start)
int.end <- pretty(int.end)

# Insert ":" separator between hours and minutes

int.start <- paste(substr(int.start, 1, 2), ":", substr(int.start, 3, 4), sep="")
int.end <- paste(substr(int.end, 1, 2), ":", substr(int.end, 3, 4), sep="")

# Prepare interval string for printing, with square brackets and dash sign

int.string <- paste("[", int.start, "-", int.end, "]", sep="")
noquote("Interval with highest average step count:")
int.string
```

Imputing Missing Values  
-----------------------  

In this section, we will perform the following:  

- Calculate how many rows have NA in the $steps column
- Devise a strategy for filling the NA values
- Build a new data frame with all the missing values filled in
- Plot histogram and calculate mean and median for total daily steps in new dataset
- Compare these mean and median values to those without the imputed values

```{r}

# Number of rows with missing values for steps

nrow.miss <- nrow(activ.dat) - nrow(activ.dat.no.na)
noquote("The number of rows with missing values for steps is: ")
nrow.miss
```

Strategies for imputing the missing values:  

NOTE: The analysis that evaluates different strategies below can be skipped.  
NOTE: I played around a little to assess the per-day-mean approach to replace NAs,   
vs the per-interval-mean approach to replace NAs with means from same intervals. 
  
That part wasn't required to be laid out explicitly.  
If you want to skip to the punch line, I ended up replacing the NA values  
with daily averages from the same day where the NA values fell.  
  
My comments below say as much, but after some elements of the analysis...  
  
[Analysis Part (can be skipped if bored)]
-----------------------------------------  
  
We check the standard deviations of two sets of distributions.  
The first is the distribution of total daily steps over the days of the study.  
The second is the set of individual distributions of intra-interval steps,  
for each interval, and averaged all of the days of the study.  
  
Standard deviation of per-interval values calculated from daily sums:  
  
```{r}
daily.steps.per.interv <- daily.sums / nlevels(as.factor(activ.dat$interval))
sd.from.daily.sums <- sd(daily.steps.per.interv)
sd.from.daily.sums
```
  
With a standard deviation of 14.8, and with a mean and median for daily steps  
being very close to one another at 10766 for total daily steps, and therefore  
at 10766/288 = 37.38 for values we'd plug into the missing slots, we would be  
using mean/median from a fairly symmetric distribution and with a coefficient  
of variation (standard deviation divided by mean) of about 0.4.  
  
So the approach above would be a pretty robust approach of imputing values.  
  
If, as an alternative, we used the data from the individual 5-minute intervals,  
and filled a missing value in one such interval with the median or the mean of  
that particular interval where the missing value lies, we would want to check  
how that might perform.  
  
Let's calculate the medians of the per-interval distributions, compare these  
to the means of the per-interval distributions, and also calculate each of  
the per-interval standard deviations, to have some sense for the robustness  
of this second approach.  
  
```{r}

# Medians within Intervals

interval.meds <- tapply(activ.dat.no.na$steps, activ.dat.no.na$interval, median, simplify=T)
interval.meds

# Standard Deviations within Intervals

interval.sds <- tapply(activ.dat.no.na$steps, activ.dat.no.na$interval, sd, simplify=T)
interval.sds
```

We see that many of the medians are zero-valued where the means are not.  
We also see that the standard deviations are very small for some intervals  
(which would have been good if we imputed from means) but much larger for  
some other intervals, particularly those where the modes of the steps are.  
  
So, all in all, it might be more robust and and slightly simpler to stick  
with an imputation method that fills the missing values with the values  
from the daily sums, divided by the number of daily intervals, i.e., 288.  
  
[Resume here if skipped the optional analysis above]
----------------------------------------------------
  
We now create the new data set with missing values imputed using the mean  
from the daily sums, divided by the number of daily intervals (288).  
  
```{r}
activ.newdat <- activ.dat
activ.newdat[is.na(activ.dat$steps), 1] <- daily.mean / nlevels(as.factor(activ.newdat$interval))
```
  
We now plot the histogram on this new dataset with imputed values for steps:  
  
```{r}
new.daily.sums <- tapply(activ.newdat$steps, activ.newdat$date, sum, simplify=T)

# Remove the NA values

new.daily.sums <- subset(new.daily.sums, !is.na(new.daily.sums))

# Plot histogram of daily steps

noquote("Plotting Histogram of Daily Steps With Imputed Data:")
hist(new.daily.sums)

# Plot histogram again into a png file

png("./figures/daily-steps-histogram-with-imputed-data.png")
hist(new.daily.sums)
dev.off()
```
  
Calculate the new values for mean and median of daily sums of steps.  
These are computed on the new dataset with the imputed values where  
there was missing data in the original dataset.  
  
```{r}
# Calculate new mean and print it

new.daily.mean <- mean(new.daily.sums)
noquote("New Daily Step Mean: ")
new.daily.mean

# Calculate new median and print it

new.daily.median <-median(new.daily.sums)
noquote("New Daily Step Median: ")
new.daily.median
```
  
We make the following observations on the new dataset:  
  
While the initial values for mean and median of total daily steps were already  
very close at 10,766.19 and 10,765, respectively, the new values for mean and  
median are now both 10,766.19. So, by imputing a common value on a fairly large  
number of missing NA entries, we have made the distribution even more centered  
around a common mean, median, mode.  
  
In terms of summary statistics, however, the imputation has barely had any impact  
on the estimates for total daily steps, when using mean or median for estimating.  
  
Activity Patterns Between Week Days and Week Ends
-------------------------------------------------
  
We create a new factor variable in the data frame, called daytype.  
The factor has two values: "weekday" and "weekend".  
  
```{r}

# First turn the date variable into a Date type

activ.newdat$date <- as.Date(activ.newdat$date)

# Now create new factor with type of day in the week, called daytype
# It will take one of two values (levels): "weekday" or "weekend"

activ.newdat$daytype <- "weekday"
activ.newdat$daytype[weekdays(activ.newdat$date) %in% c("Saturday", "Sunday")] <- "weekend"

# Make it a factor

activ.newdat$daytype <- as.factor(activ.newdat$daytype)
```
  
Now plot the steps time series over the 5-min intervals, factored by daytype:  
  
```{r fig.width=6, fig.height=8}

# Calculate vector of means within each interval, averaged over weekdays in the dataset

activ.newdat$interval <- as.factor(activ.newdat$interval)

# Split data frame into 2, one for weekdays and one for weekends

weekday.dat <- activ.newdat[activ.newdat$daytype == "weekday",]
weekend.dat <- activ.newdat[activ.newdat$daytype == "weekend",]

# Generate the 2 vectors tha will be on the y-axis in the 2 plots

weekday.interval.means <- tapply(weekday.dat$steps, weekday.dat$interval, mean, simplify=T)
weekend.interval.means <- tapply(weekend.dat$steps, weekend.dat$interval, mean, simplify=T)

# Split the screen vertically

par(mfrow = c(2,1))

plot(levels(weekend.dat$interval), weekend.interval.means, xaxt = "n", xlab="Interval", ylab="Number of Steps", main="weekend", type="l")

# Create a vector of labels to mark the ticks on the x-axis
# Put these labels at the computed locations for tick marks
# Parameter 1 describes the x-axis in axis-related functions

x.labels <- c("00:00", "05:00", "10:00", "15:00", "20:00")
axis(1, at=axTicks(1), labels=x.labels)

plot(levels(weekday.dat$interval), weekday.interval.means, xaxt = "n", xlab="Interval", ylab="Number of Steps", main="weekday", type="l")

axis(1, at=axTicks(1), labels=x.labels)

# Now plot again into a png file within the ./figures directory

png("./figures/stacked-line-plots-per-5min-interval-weekday-weekend.png")

par(mfrow = c(2,1))

plot(levels(weekend.dat$interval), weekend.interval.means, xaxt = "n", xlab="Interval", ylab="Number of Steps", main="weekend", type="l")
axis(1, at=axTicks(1), labels=x.labels)

plot(levels(weekday.dat$interval), weekday.interval.means, xaxt = "n", xlab="Interval", ylab="Number of Steps", main="weekday", type="l")
axis(1, at=axTicks(1), labels=x.labels)

dev.off()
```

